---
title: "Predicting Employee Attrition"
subtitle: "An Analysis of Key Factors Leading to Employee Turnover"
author: "Andrew Yule"
format: html
editor_options: 
  chunk_output_type: console
---
## Introduction
The specific goals of this analysis are to:  

1.  Identify the top three factors that contribute to attrition  

2.  Learn about any job role specific trends that may exist in the data set (e.g., “Data Scientists have the highest job satisfaction”).  

3.  Provide any other interesting trends and observations from your analysis.  

4.  Build a model to predict attrition

## Executive Summary
To be filled in

## Libraries and Data Loading
```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
library(tidyverse)
library(class)
library(caret)
library(e1071)

theme_set(theme_bw())
```

```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
employeeData = read_csv("https://raw.githubusercontent.com/ayule89/SMU_DDS_CaseStudy2/main/Data/CaseStudy2-data.csv")
```

## Identifying the top three factors that contribute to attrition

#### How many of the 870 employees in the data actually left?  
There were 140 out of 870 employees (16%) that left.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |> count(Attrition)
```

### Screening all variables to visually investigate which parameters are the highest indicators of attrition  
There are a number of variables that could attribute to employee attrition. We will first visually screen each of the variables to look for the top 3 factors. The full list of variables includes: 

- Age  
- Business Travel  
- Daily Rate  
- Department  
- Distance From Home  
- Education  
- Education Field  
- Environment Satisfaction  
- Gender  
- Hourly Rate  
- Job Involvement  
- Job Level  
- Job Role  
- Job Satisfaction  
- Marital Status  
- Monthly Income  
- Monthly Rate  
- Number of Previous Companies Worked For  
- Over Time  
- Percent Salary Hike  
- Performance Rating  
- Relationship Satisfaction  
- Stock Options  
- Total Working Years  
- Training Times Last Year  
- Work Life Balance  
- Years At Company  
- Years In Current Role  
- Years Since Last Promotion  
- Years With Current Manager  

### Attrition based on age  
Attrition looks to be slightly higher in people around 20 years of age.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  mutate(Age = round(Age/5)*5) |>
  group_by(Age) |>
  count(Attrition) |>
  ggplot(aes(x = Age, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on business travel  
There is an increase in attrition rate as travel increases
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(BusinessTravel) |>
  count(Attrition) |>
  ggplot(aes(x = BusinessTravel, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on daily rate?  
Attrition looks to be slightly higher around the 400 dollar group, but consistent everywhere else
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = DailyRate)) +
  geom_histogram()

employeeData |>
  mutate(DailyRate = round(DailyRate/400)*400) |>
  group_by(DailyRate) |>
  count(Attrition) |>
  ggplot(aes(x = DailyRate, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on department?  
Nothing significant as far various departments having higher / lower attrition
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(Department) |>
  count(Attrition) |>
  ggplot(aes(x = Department, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on commute?  
Attrition looks to be significantly higher for people around 15 miles of distance traveled or higher
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = DistanceFromHome)) +
  geom_histogram()

employeeData |>
  mutate(DistanceFromHome = round(DistanceFromHome/5)*5) |>
  group_by(DistanceFromHome) |>
  count(Attrition) |>
  ggplot(aes(x = DistanceFromHome, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on education  
Attrition decreases with education level consistently
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(Education) |>
  count(Attrition) |>
  ggplot(aes(x = Education, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on education field  
Attrition is slightly higher in HR followed by Technical Degree, then marketing
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(EducationField) |>
  count(Attrition) |>
  ggplot(aes(x = EducationField, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on environment satisfaction  
Low environment satisfaction has slightly higher attrition, otherwise 2-4 are ~equal
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(EnvironmentSatisfaction) |>
  count(Attrition) |>
  ggplot(aes(x = EnvironmentSatisfaction, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on gender  
The breakdown in preference between males and females looks to be insignificant
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(Gender) |>
  count(Attrition) |>
  ggplot(aes(x = Gender, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on hourly rate  
No significant trends in hourly rate and attrition
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = HourlyRate)) +
  geom_histogram()

employeeData |>
  mutate(HourlyRate = round(DailyRate/200)*200) |>
  group_by(HourlyRate) |>
  count(Attrition) |>
  ggplot(aes(x = HourlyRate, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on job involvement  
High correlation with job involvement and attrition
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(JobInvolvement) |>
  count(Attrition) |>
  ggplot(aes(x = JobInvolvement, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on job level  
1 job level has higher attrition, but other than that no obvious trends
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(JobLevel) |>
  count(Attrition) |>
  ggplot(aes(x = JobLevel, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on job role  
Significantly higher attrition in sales representative
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(JobRole) |>
  count(Attrition) |>
  ggplot(aes(x = JobRole, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on job satisfaction  
1 job level has higher attrition decreasing as job satisfaction increases
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(JobSatisfaction) |>
  count(Attrition) |>
  ggplot(aes(x = JobSatisfaction, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on marital status  
Higher attrition for single and decreasing for married and further for divorced
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(MaritalStatus) |>
  count(Attrition) |>
  ggplot(aes(x = MaritalStatus, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on monthly income  
Highest attrition in less than 5000 monthly income
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = MonthlyIncome)) +
  geom_histogram()

employeeData |>
  mutate(MonthlyIncome = round(MonthlyIncome/5000)*5000) |>
  group_by(MonthlyIncome) |>
  count(Attrition) |>
  ggplot(aes(x = MonthlyIncome, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on monthly rate  
Slightly highest around 10000 decreasing lower and higher
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = MonthlyRate)) +
  geom_histogram()

employeeData |>
  mutate(MonthlyRate = round(MonthlyRate/5000)*5000) |>
  group_by(MonthlyRate) |>
  count(Attrition) |>
  ggplot(aes(x = MonthlyRate, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on number of previous companies  
Highest for people who have worked at 5 or more companies, meaning people might have a tendency to change jobs a lot. There's also a slight uptick at low which may indicate individuals that are prone to leave frequently.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(NumCompaniesWorked) |>
  count(Attrition) |>
  ggplot(aes(x = NumCompaniesWorked, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on working over time  
Highest attrition in people working overtime
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(OverTime) |>
  count(Attrition) |>
  ggplot(aes(x = OverTime, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on percent salary hike  
No real obvious trends. Slightly higher around the 22-24% salary increase range
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(PercentSalaryHike) |>
  count(Attrition) |>
  ggplot(aes(x = PercentSalaryHike, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on performance rating  
Nothing significant in performance rating
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(PerformanceRating) |>
  count(Attrition) |>
  ggplot(aes(x = PerformanceRating, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on relationship satisfaction  
Slightly higher in lower relationship satisfaction
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(RelationshipSatisfaction) |>
  count(Attrition) |>
  ggplot(aes(x = RelationshipSatisfaction, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on stock option level  
Highest among 0 and 3 (lower at 1 and 2)
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(StockOptionLevel) |>
  count(Attrition) |>
  ggplot(aes(x = StockOptionLevel, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on total working years  
Attrition is highest in people during their first 5 years and then levels off. At 40 years people begin retiring.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = TotalWorkingYears)) +
  geom_histogram()

employeeData |>
  mutate(TotalWorkingYears = round(TotalWorkingYears/5)*5) |>
  group_by(TotalWorkingYears) |>
  count(Attrition) |>
  ggplot(aes(x = TotalWorkingYears, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on training times  
No noticeable patterns
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(TrainingTimesLastYear) |>
  count(Attrition) |>
  ggplot(aes(x = TrainingTimesLastYear, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on work life balance  
Significant trend in attrition with lower work life balance and decreasing as work life balance improves
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(WorkLifeBalance) |>
  count(Attrition) |>
  ggplot(aes(x = WorkLifeBalance, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on years at company  
Slightly higher for people in their first 5 years. People over 40 are retiring.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = YearsAtCompany)) +
  geom_histogram()

employeeData |>
  mutate(YearsAtCompany = round(YearsAtCompany/5)*5) |>
  group_by(YearsAtCompany) |>
  count(Attrition) |>
  ggplot(aes(x = YearsAtCompany, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on years at company  
Slightly higher for people in their first 5 years. People worked over 20 years in the current role show no attrition.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = YearsInCurrentRole)) +
  geom_histogram()

employeeData |>
  mutate(YearsInCurrentRole = round(YearsInCurrentRole/5)*5) |>
  group_by(YearsInCurrentRole) |>
  count(Attrition) |>
  ggplot(aes(x = YearsInCurrentRole, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on years since last promotion  
Slightly higher attrition for people over 15 years since their last promotion but otherwise no trends
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = YearsSinceLastPromotion)) +
  geom_histogram()

employeeData |>
  mutate(YearsSinceLastPromotion = round(YearsSinceLastPromotion/5)*5) |>
  group_by(YearsSinceLastPromotion) |>
  count(Attrition) |>
  ggplot(aes(x = YearsSinceLastPromotion, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```

#### Attrition based on years with current manager  
Longer time spent with current manager the less likely there is for attrition
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  ggplot(aes(x = YearsWithCurrManager)) +
  geom_histogram()

employeeData |>
  mutate(YearsWithCurrManager = round(YearsWithCurrManager/5)*5) |>
  group_by(YearsWithCurrManager) |>
  count(Attrition) |>
  ggplot(aes(x = YearsWithCurrManager, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```
