---
title: "Factors Leading to Attrition"
author: "Andrew Yule"
format: html
editor_options: 
  chunk_output_type: console
---
## Introduction
The specific goals of this analysis are to:  
1.  Identify the top three factors that contribute to attrition (note, there may or may not be a need to create derived attributes/variables/features)  
2.  Learn about any job role specific trends that may exist in the data set (e.g., “Data Scientists have the highest job satisfaction”).  
3.  Provide any other interesting trends and observations from your analysis.  
4.  Build a model to predict attrition

## Executive Summary
To be filled in

## Libraries and Data Loading
```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
library(tidyverse)
library(class)
library(caret)
library(e1071)

theme_set(theme_bw())
```

```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
employeeData = read_csv("/Users/andrewyule/Dropbox/Personal/SMU MSDS/03 - Doing Data Science/SMU_DDS_CaseStudy2/Data/CaseStudy2-data.csv")
```

## Identifying the top three factors that contribute to attrition

To start, how many of the 870 employees in the data actually left?  
There is around 140 out of 870 employees that left.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |> count(Attrition)
```

There are a number of variables that could attribute to employee attrition. We will try and quickly screen all of them to look for top factors to build a model around. Among the possible list are:  
- Age  
- Business Travel  
- Daily Rate  
- Department  
- Distance From Home  
- Education  
- Education Field  
- Environment Satisfaction  
- Gender  
- Hourly Rate  
- Job Involvement  
- Job Level  
- Job Role  
- Job Satisfaction  
- Marital Status  
- Monthly Income  
- Monthly Rate  
- Number of Previous Companies Worked For  
- Over Time  
- Percent Salary Hike  
- Performance Rating  
- Relationship Satisfaction  
- Stock Options  
- Total Working Years  
- Training Times Last Year  
- Work Life Balance  
- Years At Company  
- Years In Current Role  
- Years Since Last Promotion  
- Years With Current Manager  

Is there any trends in attrition based on age grouped by 5 years  
X
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  mutate(Age = round(Age/5)*5) |>
  group_by(Age) |>
  count(Attrition) |>
  ggplot(aes(x = Age, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```


### Is there any preference for attrition between males and females?
The breakdown in preference between males and females looks to be insignificant
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
employeeData |>
  group_by(Gender) |>
  count(Attrition) |>
  ggplot(aes(x = Gender, y = n, fill = Attrition)) +
  geom_col(position = "fill")
```
