---
title: "Predicting Employee Attrition"
subtitle: "An Analysis of Key Factors Leading to Employee Turnover"
author: "Andrew Yule"
format: html
editor_options: 
  chunk_output_type: console
---
## Introduction
The specific goals of this analysis are to:  

1.  Identify the top three factors that contribute to attrition  

2.  Learn about any job role specific trends that may exist in the data set (e.g., “Data Scientists have the highest job satisfaction”).  

3.  Provide any other interesting trends and observations from your analysis.  

4.  Build models to predict attrition and salaries

## Executive Summary
To be filled in

## Libraries and Data Loading
```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
#| message: false
library(tidyverse)
library(gridExtra)
library(reshape2)
library(class)
library(caret)
library(e1071)

theme_set(theme_bw())
```

```{r}
#| echo: true
#| output: false
#| warning: false
#| error: false
#| message: false
employeeData = read_csv("https://raw.githubusercontent.com/ayule89/SMU_DDS_CaseStudy2/main/Data/CaseStudy2-data.csv")
```

## 1. Identifying the top three factors that contribute to attrition

#### How many of the 870 employees in the data actually left?  
There were 140 out of 870 employees (16%) that left.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeData |> count(Attrition)
```

### Screening all variables to visually investigate which parameters are the highest indicators of attrition  
There are a number of variables that could attribute to employee attrition. We will first visually screen each of the variables to look for potential leading indicators. The full list of variables includes:
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
colnames(employeeData)
```

Before investigating attrition across each variable, some of the numerical scales will need to put into various bins to make analysis easier. In order to determine appropriate bins, we'll first plot histograms of the numerical variables, which include: Age, DailyRate, DistanceFromHome, HourlyRate, MonthlyIncome, MonthlyRate, TotalWorkingYears, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearsWithCurrManager.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeData |>
  select(Age, DailyRate, DistanceFromHome, HourlyRate, MonthlyIncome, MonthlyRate, TotalWorkingYears, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearsWithCurrManager) |>
  pivot_longer(cols = everything(), names_to = "column", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram() +
  facet_wrap(~column, scales = 'free', ncol = 3)
```

Based on visually inspecting the histograms, the following bins for various numerical variables will be used below. In addition, we'll change the Attrition values from strings to doubles with Yes/No being 1/0 respectively.

- Age (5 years)  
- Daily rate (400 dollars)  
- Distance from home (5 miles)  
- Hourly rate (20 dollars)  
- Monthly income (5,000 dollars)  
- Monthly rate (5,000 dollars)
- Total working years (5 years)  
- Years at company (5 years)  
- Years in current role (5 years)  
- Years since last promotion (5 years)  
- Years under current manager (5 years)

```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeDataBinned = employeeData |>
  mutate(
    Age = round(Age/5)*5, 
    DailyRate = round(DailyRate/400)*400,
    DistanceFromHome = round(DistanceFromHome/5)*5,
    HourlyRate = round(HourlyRate/20)*20,
    MonthlyIncome = round(MonthlyIncome/5000)*5000,
    MonthlyRate = round(MonthlyRate/5000)*5000,
    TotalWorkingYears = round(TotalWorkingYears/5)*5,
    YearsAtCompany = round(YearsAtCompany/5)*5,
    YearsInCurrentRole = round(YearsInCurrentRole/5)*5,
    YearsSinceLastPromotion = round(YearsSinceLastPromotion/5)*5,
    YearsWithCurrManager = round(YearsWithCurrManager/5)*5)
```

Now we'll go through and create plots of each variable showing the breakdown between attrition to visually look for any strong relationships
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
#| fig-width: 12
#| fig-height: 15
# Helper function that creates a plot showing the breakdown in attrition for a given variable across it's possible values
findAttritionBreakdown = function(df, colName) {
  df |>
    group_by(!!sym(colName)) |>
    count(Attrition) |>
    ungroup() |>
    ggplot(aes(x = !!sym(colName), y = n, fill = Attrition)) +
    geom_col(position = "fill") +
    coord_cartesian(ylim = c(0, 0.55))
}
# Test the function
findAttritionBreakdown(employeeDataBinned, "DistanceFromHome")

cols = colnames(employeeData)
# Drop columns that we don't want to include in the analysis (ID, Attrition, EmployeeCount, EmployeeNumber, Over18, StandardHours) 
cols = cols[c(-1, -3, -10, -11, -23, -28)]

# Create a list of plots across all of the columns using the map function from the purr package
plots = map(cols, function(col) findAttritionBreakdown(employeeDataBinned, col))
# Display plots on one page
grid.arrange(grobs = plots, ncol = 3)
```

Based on examining the plots above, the following key variables were identified as likely having higher impacts on employee attrition:

- Age  
- BusinessTravel  
- DistanceFromHome  
- Education  
- EnvironmentSatisfaction  
- JobInvolvment  
- JobRole  
- JobSatisfaction  
- MaritalStatus  
- MonthlyIncome  
- NumCompaniesWorked  
- OverTime  
- StockOptionLevel  
- TotalWorkingYears  
- WorkLifeBalance  
- YearsWithCurrManager  

In order to determine the top 3 factors quantitatively, we will calculate the Pseudo-R2 value for each predictor. This is based on a similar analysis that can be found [here](https://heap.io/blog/pseudo-r2-a-metric-for-quantifying-interestingness).

The main idea is that we first start with a Null model, or a baseline model, that can be used to determine the likelihood of attrition. Once that is established, then additional factors can be evaluated against this Null model using their likelihood to look for improvements or increases in the likelihood. In this case, because the likelihoods can be quite large, loglikelihood values will be used.

The Null model in this case will simply be the LogLikelihood of attrition in the entire data set or XX
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
calcLogLikelihoodData = function(data){
  no = select(filter(data, Attrition == 0), n)
  yes = select(filter(data, Attrition == 1), n)
  if(dim(yes)[1] == 0 || dim(no)[1] == 0) {
    return(tibble(n = c(0)))
  }
  else{
    p = yes/(yes + no)
    log(p^yes * (1-p)^no)  
  }
}

employeeDataBinnedWithQuantitativeAttrition = employeeDataBinned |>
  mutate(Attrition = ifelse(Attrition == "Yes", 1, 0))

nullModel = calcLogLikelihoodData(count(employeeDataBinnedWithQuantitativeAttrition, Attrition))

calcPseudoRSquared = function(var){
  employeeDataBinnedWithQuantitativeAttrition |>
    group_by(!!sym(var)) |>
    count(Attrition) |>
    group_modify(function(group, blah) calcLogLikelihoodData(select(group, Attrition, n))) |>
    ungroup() |>
    select(n) |>
    sum() |>
    (function(ll) (1-ll/nullModel$n)*100)()
}
#Test
calcPseudoRSquared("Age")

# Calculate Pseudo-R2 for all variables
pseudoRSquared = tibble("Variable" = cols) |>
  rowwise() |>
  mutate(PseudoRSquared = calcPseudoRSquared(Variable))

pseudoRSquared |>
  ggplot(aes(x = PseudoRSquared, y = fct_reorder(Variable, PseudoRSquared))) +
  geom_col() +
  labs(x = "Pseudo-RSquared", y = NULL, title = "Pseudo-RSquared Values for Each Variable")
```

Based on the analysis above, the top 3 factors influencing employee attrition are:

1. Job Role  
2. Stock Option Level  
3. Over Time

The next 3 factors following the top 3 are:

1. Total Working Years  
2. Years At Company  
3. Monthly Income

## 2. Learn about any job role specific trends that may exist in the data set

We know that Job Role is our number one predictor of employee attrition. Looking closer at various reported job roles, it is clear that sales type positions see much larger amounts of turnover while managing and director type roles see very little.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeData |>
  group_by(JobRole) |>
  summarize(AttritionPct = 100*sum(ifelse(Attrition == "No", 0, 1))/n()) |>
  ggplot(aes(x = AttritionPct, y = reorder(JobRole, AttritionPct))) +
  geom_col() +
  labs(title = "Attrition Percentage Across Various Job Roles", x = "Attrition Percentage", y = NULL)
```

Knowing that stock options, over time, and monthly income are key also key indicators of attrition, we can use those to better understand why sales positions see higher attrition.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
# Looking at stock options
employeeDataBinnedWithQuantitativeAttrition |>
  mutate(JobRole = factor(JobRole, levels = c("Sales Representative", "Human Resources", "Laboratory Technician", "Research Scientist", "Sales Executive", "Healthcare Representative", "Manager", "Manufacturing Director", "Research Director"))) |>
  group_by(JobRole) |>
  count(StockOptionLevel) |>
  ungroup() |>
  ggplot(aes(x = JobRole, y = n, fill = StockOptionLevel)) +
  geom_col(position = "fill")

# Looking at over time
employeeDataBinnedWithQuantitativeAttrition |>
  mutate(JobRole = factor(JobRole, levels = c("Sales Representative", "Human Resources", "Laboratory Technician", "Research Scientist", "Sales Executive", "Healthcare Representative", "Manager", "Manufacturing Director", "Research Director"))) |>
  group_by(JobRole) |>
  count(OverTime) |>
  ungroup() |>
  ggplot(aes(x = JobRole, y = n, fill = OverTime)) +
  geom_col(position = "fill")

# Looking at monthly income
employeeData |>
  mutate(JobRole = factor(JobRole, levels = c("Sales Representative", "Human Resources", "Laboratory Technician", "Research Scientist", "Sales Executive", "Healthcare Representative", "Manager", "Manufacturing Director", "Research Director"))) |>
  ggplot(aes(x = JobRole, y = MonthlyIncome)) +
  geom_boxplot()
```

Analyzing Sales Representatives individually we see that they rank very highly in all of the categories that contribute the most to attrition. They have a proportionally smaller amount of stock options generally, they work slightly more overtime than most other positions with the exception of research scientist, and they are the lowest paid.

## 3. Provide any other interesting trends and observations from your analysis

Let's look at some of the quantitative variables to look for any interesting trends
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeData |>
  select(Age, DistanceFromHome, Education, JobInvolvement, JobLevel, JobSatisfaction, MonthlyIncome, NumCompaniesWorked, PercentSalaryHike, PerformanceRating, RelationshipSatisfaction, TotalWorkingYears, TrainingTimesLastYear, WorkLifeBalance, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearsWithCurrManager) |>
  cor() |>
  round(2) |>
  melt() |>
  ggplot(aes(x=Var1, y=Var2, fill=value)) +
  geom_tile() +
  theme(axis.text.x = element_text(angle = 90))
```

From the above correlation heat map, we can make the following observations:

- Years at company, years in current role, years since last promotion, and years with current manager are all closely related as they all relate to how long a particular employee has been at the current job  
- High performance rating is correlated with higher salary hikes  
- There were no strong correlations for work life balance which could mean there are other non-numerical indicators for work life balance, or it depends more on the individual making it difficult to predict

## 4. Build models to predict attrition and salary

To predict attrition, a naive Bayes model was used with the goal of reaching the highest accuracy while preserving at least 60% for sensitivity and specificity. After cross-validating the model using a 75/25 train/test split and utlizing 16 of the top 30 predicting variables, the respective achievable accuracy, sensitivity, and specificity was found 80%, 84%, and 61%.
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false
employeeDataForPrediction = employeeData |>
  select(-ID, -EmployeeCount, -EmployeeNumber, -Over18, -StandardHours)

set.seed(1234)
evaluateModel = function() {
  l = length(employeeDataForPrediction$Attrition)
  indicies = sample(seq(1:l), round(0.75*l))
  train = employeeDataForPrediction[indicies,]
  test = employeeDataForPrediction[-indicies,]
  
  model = naiveBayes(Attrition ~ JobRole + StockOptionLevel + OverTime + TotalWorkingYears + YearsAtCompany + MonthlyIncome + JobLevel + Age + MaritalStatus + JobInvolvement + YearsInCurrentRole + NumCompaniesWorked + YearsWithCurrManager + PercentSalaryHike + WorkLifeBalance + JobSatisfaction, train)
  predictions = predict(model, select(test, JobRole, StockOptionLevel, OverTime, TotalWorkingYears, YearsAtCompany, MonthlyIncome, JobLevel, Age, MaritalStatus, JobInvolvement, YearsInCurrentRole, NumCompaniesWorked, YearsWithCurrManager, PercentSalaryHike, WorkLifeBalance, JobSatisfaction))
    
  results = confusionMatrix(table(predictions, test$Attrition))
  acc = as.numeric(results$overall[1])
  sen = sensitivity(table(predictions, test$Attrition))
  spc = specificity(table(predictions, test$Attrition))
  
  return(tibble("Accuracy" = c(acc), "Sensitivity" = c(sen), "Specificity " = c(spc)))
}

results = tibble("Accuracy" = c(NULL), "Sensitivity" = c(NULL), "Specificity " = c(NULL))
for (variable in 1:200) {
  results = rbind(results, evaluateModel())
}

results |>
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(x = Variable, y = Value)) +
  geom_boxplot()
```

To predict salaries, a linear regression model was used
```{r}
#| echo: true
#| output: true
#| warning: false
#| error: false
#| message: false

```



